---
title: "Supplemental Materials"
subtitle: "The current state of knowledge and data on ectomycorrhizal fungi in Canada"
author: "Eckert, Qin, Kivlin, Shelton, Yusta Belsham, Fischer, Karst & Pither"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    df-print: kable
execute:
  echo: false
  warning: false
  message: false
editor: source
---

```{r setup}
#| include: false
# Source the shared pipeline setup: loads packages, the `paths` list, helper
# functions, and (if it exists) the primary EcM dataset `emf`.
source(here::here("scripts", "00_setup.R"))

# Small helper for rendering data-frame tables with a caption.
show_table <- function(x, caption = NULL, ...) {
  knitr::kable(x, caption = caption, format = "html", escape = FALSE, ...)
}

# Convenience: absolute path to a figure by file name.
fig <- function(name) here::here("figures", name)
```

# Supplemental methods

These methods provide the detailed procedures underlying every analysis and
result reported in the main text. They are organized to follow the logic of the
pipeline: first the assembly of the Canadian ectomycorrhizal (EcM) fungal
dataset and its supporting reference layers, then the methods for each of the
seven biodiversity shortfalls in the order they appear in the manuscript
(Linnean, Wallacean, Prestonian, Darwinian, Raunkiæran, Hutchinsonian,
Eltonian). All analyses were conducted in R (v4.5.2; R Core Team 2026); the
complete, ordered code base is described in `scripts/README.md`.

## Assembly of the Canadian EcM fungal dataset

### Data sources

We compiled EcM fungal records for Canada from two publicly accessible sequence
sources: version 5 of the GlobalFungi database (Větrovský et al. 2020) and NCBI
GenBank (accessed 29 June 2026). GlobalFungi is a curated, purpose-built
database of community-survey amplicon data, consistently processed but
aggregated from studies using varied sequencing platforms and protocols, and
primarily derived from soil samples. GenBank is a general-purpose sequence
repository comprising records deposited from diverse studies and sampling
efforts, including targeted root sampling. Unlike GlobalFungi records, GenBank
records lack per-sample read counts or a consistent sampling frame, so they
cannot be used to estimate facets of diversity that require a defined sampling
unit.

### UNITE reference database and SH assignment

We assigned fungal species hypothesis (SH) codes and taxonomy using the UNITE
general FASTA release (build 4 April 2024, UNITE v10.0, DOI
10.15156/BIO/2959332; dynamic clustering, full-length ITS; Nilsson et al. 2019).
This release provides one representative sequence per SH code, each with a
single unambiguous taxonomy string in its header, which we parsed into a lookup
table mapping each SH code to kingdom through species. Because each SH is
represented by exactly one sequence, that representative's taxonomy is taken
directly as the SH-level taxonomy, with no consensus step. We used the
full-length (`_dev`) variant, which retains flanking rRNA regions and is
appropriate for matching full-length GenBank query sequences.

We pinned the pipeline to the 4 April 2024 build. UNITE SH identifiers are
renumbered between dated builds of the same major version, so taxonomy must be
decoded against the same build that assigned the SH codes; the 4 April 2024
build resolves taxonomy for >99% of GlobalFungi's pre-assigned Canadian SH
codes, and GenBank SH codes are assigned in-house against this same build (see
below), so both sources are internally consistent.

### GlobalFungi records

We downloaded the GlobalFungi v5 sample-metadata file and the combined
ITS1/ITS2 SH abundance matrix (samples × SH codes). We retained Canadian samples
by filtering for country = Canada; barcoding region in {ITS2, ITSboth};
non-manipulated conditions (`manipulated == "NO"`); and exclusion of substrate
types irrelevant to EcM fungi (shoot, air, water, sediment). We pivoted the
resulting wide abundance matrix to long format and removed zero-abundance
records, then joined UNITE taxonomy by SH code. Because GlobalFungi's SH codes
are pre-assigned to UNITE, no additional sequence-similarity searching was
required for this source. We confirmed that the GlobalFungi dataset already
included all 12 Canadian samples from the Global Soil Mycobiome Consortium
(GSMc; Tedersoo et al. 2021), so no independent GSMc ingestion was needed.

### GenBank records

We searched GenBank (nuccore) for fungal ITS sequences from Canada using the
query `"Canada"[Country] AND "Fungi"[Organism] AND ("internal transcribed
spacer"[All Fields] OR "ITS"[All Fields])`, with no date cutoff and no
lifestyle keyword, so the search returns all matching fungal ITS records
regardless of guild. We retrieved FASTA sequences and record metadata
(coordinates, collection date, host, isolation source) in batches through the
NCBI Entrez API (`rentrez`; Winter 2017).

We assigned sequences to UNITE SH codes with `vsearch` v2.30.4 (Rognes et al.
2016) at 98.5% global-alignment identity, retaining only the top hit per query.
This threshold matches GlobalFungi's documented SH-assignment criterion,
ensuring both sources apply the same membership rule. Taxonomy was read directly
from the matched SH representative. Host taxon was taken from the structured
GenBank `host` qualifier where available, with a fallback regex extraction of
binomial names from `isolation_source`. The raw extracted string is preserved
as `host_taxon_raw`; the canonicalized `host_taxon` used by all downstream
analyses is derived from it via the shared `canonicalize_host()` helper (strip
surrounding punctuation, collapse whitespace, truncate to the first two tokens,
apply a curated typo-correction lookup, case-normalize, and reject entries that
fail a Latin-name pattern check).

We applied two post-assignment quality filters to improve sequence quality and
ITS-marker consistency with the GlobalFungi component: we excluded sequences
shorter than 200 bp, and records whose `vsearch` alignment against the UNITE
reference was shorter than 200 bp. The second filter acts as a practical proxy
for ITS-region consistency: very short alignments are characteristic of
ITS1-only sequences, and screening them out gives marker-type consistency with
the ITS2/ITSboth GlobalFungi component. Each GenBank record was also assigned a
`canada_basis` provenance category recording the evidence for its inclusion
(structured country field, coordinates, both, or search-only).

### Combining sources, spatial validation, and EcM filtering

We combined the two datasets by row-binding, retaining source-specific columns
as `NA` for the other source. For all records with parsed coordinates we applied
a spatial-containment check against the GADM administrative boundary for Canada,
recording the result in a logical `coord_in_canada` column (TRUE = validated
inside the boundary; FALSE = coordinates present but outside; NA = no
coordinates); original coordinates are always preserved. Spatial analyses filter
on `coord_in_canada == TRUE`, while non-spatial enumerations (taxon richness,
record counts) include all records, treating a GlobalFungi `country == Canada`
label as authoritative even where coordinates fall marginally outside the
polygon. We disabled S2 spherical geometry locally around GADM predicate
operations (`sf::sf_use_s2(FALSE)`), which the GEOS engine handles despite minor
boundary-topology imperfections, and re-enabled it immediately afterward.

We determined EcM status at the genus level using FungalTraits (Põlme et al.
2020): we retained records whose UNITE-assigned genus carries
`primary_lifestyle = ectomycorrhizal` in FungalTraits (genus names lower-cased
before matching), and carried two additional FungalTraits fields — EcM lineage
and exploration type — into the final dataset. The resulting primary analysis
dataset (`data_derived/emf_canada_em_only.csv`) has one row per
(sample × SH code) detection.

### Definition of a "site"

For analyses requiring georeferenced sampling units, we binned sample
coordinates to three decimal places of latitude and longitude (~100 m at
temperate latitudes), and refer to these bins as sampling "sites". This
collapses GPS-precision noise across repeat readings at one physical location
without merging biologically distinct sites, and is applied through the shared
`add_site_id()` helper. Site binning is used for the incidence-based Chao2 and
accumulation analyses; the Wallacean occurrence counts (below) are intentionally
left unbinned so that each distinct coordinate contributes an occurrence.

## Reference data preparation

### FungalRoot host-determination rule

We determined EcM host status for plant species from the FungalRoot database
(Soudzilovskaia et al. 2020), accessed as occurrence-level Darwin Core Archive
records from GBIF (DOI 10.15468/a7ujmj). We joined the occurrence file to the
measurements extension on the mycorrhiza-type field, yielding one row per
(plant binomial, raw mycorrhizal-type label). A species is classified as
EcM-demonstrated if FungalRoot carries at least one occurrence with an
unambiguous EcM-positive label (`EcM`, `EcM, no AM colonization`, or
`ErM,EcM`). Two labels are excluded as ambiguous: `EcM, AM undetermined` (which
does not distinguish a genuine EcM record from a non-EcM record for which AM
status was not assessed) and `EcM,AM` (a dual-positive call that, on inspection,
introduced implausible hosts with no other EcM evidence). Plant names were
standardized through the GBIF taxonomic backbone
(`rgbif::name_backbone_checklist()`) before classification. This rule requires
no free parameters (no vote threshold or proportion cutoff).

### Native Canadian EcM host species

We identified native EcM host plant species occurring in Canada by querying BIEN
(Maitner et al. 2018) for the plant species with Canadian distributions,
filtering to natives via the Native Status Resolver, and intersecting that
native flora directly against the species-level EcM-demonstrated list above
(species-level selection only, with no genus-level inference and no growth-form
restriction). The resulting list
(`data_derived/ecm_native_canada_host_species.csv`) is the universe of EcM host
species used in the Hutchinsonian and Eltonian analyses.

Because BIEN's occurrence-name scrubbing, the Native Status Resolver, and the
GBIF backbone underlying FungalRoot are independent taxonomic resolvers, a
genuinely native, EcM-demonstrated species can be excluded if no single name
string clears both the native filter and the demonstrated filter
simultaneously. A confirmed example is the green/mountain alder complex
(*Alnus alnobetula* / *A. crispa* / *A. viridis*): the native filter returns
"Native" only under *Alnus alnobetula* while FungalRoot's unambiguous EcM
evidence is filed under *Alnus crispa*, so the intersection finds no shared name
and the species drops out. We flag this as a known limitation of unresolved
cross-database synonymy rather than applying a general synonym-harmonization
step (which could shift counts for many species).

### BIEN2 host ranges and richness rasters

We obtained MaxEnt-based modelled range polygons (BIEN2; Moulatlet et al. 2025)
for the native Canadian EcM host species from biendata.org, downloaded
programmatically through the REST API, reprojected to WGS84, and clipped to the
Canada boundary. We rasterized these polygons onto a 0.5° geographic grid to
produce three layers: host species richness (count of host species with a range
polygon per cell); host data richness (count of host species with at least one
Canadian EcM molecular-detection record per cell); and host data proportion
(the ratio of the two). We also flagged ecoregions containing at least one
0.5° cell with predicted EcM host habitat, restricting ecoregion-level gap
analyses to ecologically relevant areas.

## Spatial layers

We obtained Canadian national, provincial, and territorial boundaries from GADM
via `geodata`, a higher-resolution Canada boundary and major-lake polygons from
Natural Earth via `rnaturalearth`, and Canadian ecoregions (National Ecological
Framework for Canada) with their ecozone-name lookup. Layers were reprojected to
WGS84 for coordinate operations and to the Canada Albers Equal Area Conic
projection for mapping, simplified with `rmapshaper` where appropriate, and
written to `data_derived/spatial/`.

## Shortfall analyses

### Linnean shortfall — taxonomic knowledge

We quantified the Canadian EcM fungal inventory at three taxonomic levels:
unique UNITE SH codes, genera, and named species (excluding unresolved UNITE
entries ending in `_sp`). SH codes are the primary richness unit because they
provide a consistent operational taxonomic unit applicable to both sources, and
because many EcM taxa are not yet formally named. We compared observed Canadian
EcM genera to the global EcM genus inventory in FungalTraits, and quantified the
overlap between the GlobalFungi and GenBank components at the named-species SH
level using set operations (shared, GlobalFungi-only, GenBank-only). We
characterized the dataset "dark fraction" as the proportion of detected SH codes
lacking a species-level UNITE name, computed for the combined dataset and for
each source separately. This SH-level dark fraction is not directly comparable
to the OTU-level dark-diversity estimates of van Galen et al. (2025), whose
denominators include OTUs matched only to genus level; for reference we also
computed the average per-sample proportion of species-assigned OTUs among van
Galen et al.'s Canadian forest and tundra samples from their published
per-sample table.

For a within-footprint lower bound on richness, we constructed a site × SH
incidence matrix from the georeferenced Canadian GlobalFungi data (sites defined
by 3-decimal binning; singletons retained) and applied the incidence-based
Chao2 asymptotic estimator through `iNEXT::iNEXT()` (`datatype =
"incidence_freq"`, q = 0, 50 bootstrap replicates, 95% CI). GenBank records are
excluded from this estimator because opportunistic one-off accessions inflate
the singleton count (Q1) without reflecting genuine rarity. To complement the
spatial estimator, we also computed per-sample abundance-based rarefaction and
extrapolation for each Canadian GlobalFungi sample (`iNEXT`, q = 0, `datatype =
"abundance"`, 50 bootstrap replicates, endpoint = 2 × per-sample EcM read sum),
extracting observed richness, the Chao1 asymptotic estimator, and sample
completeness Ĉ. Per-sample read counts are the sum of reads mapping to
EcM-genus SHs (not total sequencing depth) and vary over several orders of
magnitude across the pooled studies; per-sample Chao1 from reads is therefore
interpreted as a standardized completeness index rather than a count of
unobserved taxa. We additionally compiled physical EcM specimen records from
GBIF (preserved and fossil specimens, Kingdom Fungi, Canada; filtered to EcM
genera via FungalTraits) as a vouchered complement to the sequence data.

### Wallacean shortfall — geographic-range knowledge

We characterized geographic sampling breadth for each EcM taxon at three levels
(SH code, genus, named species). For each taxon we counted unique sampling
occurrences as distinct 30 arc-second grid cells (~1 km²), and produced
frequency distributions of occurrences per taxon within Canada and globally. We
report the proportion of taxa known from a single occurrence, and from fewer
than 30 occurrences, both within Canada and globally, as straightforward
measures of the shortfall. The within-Canada and global occurrence counts are
drawn from the GlobalFungi dataset only, because it provides the structured
sampling frame required for meaningful occupancy comparison; the global counts
map each named Canadian species (or SH code) to all globally-known SH codes
carrying that identity and match them against the full global GlobalFungi v5 SH
abundance matrix under the same quality filters used for the Canadian data. As a
complementary view relevant to species distribution modelling, we tabulated how
many taxa meet common minimum-occurrence thresholds for modelling.

### Prestonian shortfall — temporal knowledge

We searched the BioTIME database (Dornelas et al. 2018) for time-series records
of EcM fungal communities in Canada, filtering to studies with at least one
Canadian location, at least two sampling time points, and an organism type
consistent with fungi, and cross-referencing matching studies against the EcM
dataset by taxonomy. We operationalize the Prestonian shortfall as the
percentage of named EcM fungal species present in Canada that lack any abundance
time-series data recorded anywhere on Earth.

### Darwinian shortfall — phylogenomic knowledge

We assessed genome-sequence availability for Canadian EcM genera by
cross-referencing detected genera against the JGI MycoCosm fungal-genomics
portal, matching organism names to genera after lower-case normalization and
fuzzy string matching. We operationalize the Darwinian shortfall as the
percentage of named Canadian EcM fungal species absent from the database.

### Raunkiæran shortfall — trait knowledge

We quantified trait-data availability using FungalTraits (a genus-level
compilation) for all EcM genera detected in Canada; every detected genus was
present in FungalTraits, so the shortfall is one of trait completeness rather
than genus coverage. We assessed the six FungalTraits columns carrying
information meaningful for EcM fungi, in two classes: three
morphological/structural traits derivable from taxonomic description (fruitbody
type, hymenium type, mycorrhizal exploration type) and three
ecological/interaction traits (secondary lifestyle, endophytic-interaction
capability, host specificity). The seventh potentially relevant column, growth
form, is "filamentous mycelium" for all EcM genera and was excluded as
uninformative. Coverage is the proportion of genus × trait cells with a
documented value, where "documented" means non-missing and not the literal
entry "unknown". Blanks in secondary lifestyle and endophytic-interaction
capability are ambiguous between "not assessed" and "trait absent", so coverage
for those two traits is a lower bound.

### Hutchinsonian shortfall — environmental and geographic coverage

We addressed environmental coverage at two levels. For **ecoregion coverage**,
we determined which Canadian ecoregions contain at least one EcM sampling
location, excluding ecoregions without predicted EcM host habitat (from the
BIEN2 host raster) before calculating the shortfall, and for ecoregions meeting
site thresholds we estimated ecozone-level SH richness with the same
incidence-based Chao2 estimator used for the Linnean shortfall.

For **climate-space coverage**, we characterized how completely Canada's climate
space has been sampled. We defined climate space using WorldClim 2.1
bioclimatic variables BIO1 (mean annual temperature) and BIO12 (mean annual
precipitation), aggregating the 30 arc-second rasters to ~10 arc-minutes and
masking to the Canada boundary, then binned the two-dimensional
temperature × precipitation space into a 50 × 50 grid of equal-width climate
zones. For each zone we computed coverage as the proportion of its Canadian grid
cells containing at least one EcM sample (each cell counted once), with a
coverage of zero denoting a climate present in Canada but never sampled.
Coverage was mapped in both geographic and climate space, separately for
GlobalFungi samples and for the combined GlobalFungi + GenBank dataset, against
the available-climate frequency (the number of Canadian cells per zone). The
code producing this figure was adapted from Hébert et al. (2026).

### Eltonian shortfall — biotic-interaction knowledge

We focused on EcM fungus–host plant associations, in two scopes. For the
**Canadian scope**, host information was extracted from the GenBank `host`
qualifier (with fallback from `isolation_source`) and from GlobalFungi's
`dominant_plant_species` and `other_plant_species` fields, restricted to root
samples (direct root extractions provide the strongest evidence of a
mycorrhizal association, whereas bulk-soil samples indicate only spatial
proximity). All host strings were canonicalized with the shared helper and
matched against the native Canadian EcM host species list, which serves as the
denominator of potential interactions. We built fungus × host interaction
matrices at the SH-code, named-species, and genus levels, and computed the
proportion of Canadian EcM host species with at least one documented fungal
association, the fill rate of the full host × named-fungal-species matrix, and
the proportion of observed pairs supported by only a single record. For the
**global scope**, we queried the full GlobalFungi global metadata for root
samples matching Canadian host species, and queried GenBank globally for each
Canadian EcM genus (capped at 2,000 records per genus), to establish which
Canadian hosts and fungi have any documented association anywhere. To
contextualize the sampling gap relative to host diversity, we produced a
bivariate map classifying each 0.5° grid cell by EcM host species richness
versus the proportion of those host species with EcM molecular-detection data in
Canada, shown separately for tree and non-tree hosts.

## Software

Analyses were conducted in R (v4.5.2; R Core Team 2026). Key packages: `sf`
(Pebesma 2018) for vector spatial operations; `terra` (Hijmans 2024) for
rasters; `ggplot2` (Wickham 2016) and `patchwork` (Pedersen 2024) for figures;
`vegan` (Oksanen et al. 2022) and `iNEXT` (Hsieh, Ma & Chao 2016) for
rarefaction, extrapolation, and asymptotic richness estimation; `rentrez`
(Winter 2017) for GenBank retrieval; `rgbif` (Chamberlain et al. 2024) for GBIF
access; `BIEN` (Maitner et al. 2018) for plant range and native-status data;
`geodata` and `rnaturalearth` for spatial base layers; `rmapshaper` for boundary
simplification; and `httr` for programmatic download of BIEN2 range shapefiles.
Sequence-similarity searching used `vsearch` v2.30.4 (Rognes et al. 2016), and
extraction from the GlobalFungi abundance matrix used `awk`.

# Supplemental tables and figures

The tables and figures below are generated directly from the pipeline outputs in
`data_derived/` and `figures/`. They render once the pipeline has been run; each
chunk reports if its input is not yet present.

## Table S1. Per-ecozone sample completeness

```{r table-s1-ecozone-completeness}
cov_path <- file.path(paths$out_linnean, "linnean_extrapolation_coverage.csv")
est_path <- file.path(paths$out_linnean, "linnean_extrapolation_estimators.csv")
if (file.exists(cov_path)) {
  cov_tbl <- readr::read_csv(cov_path, show_col_types = FALSE)
  if (file.exists(est_path)) {
    est_tbl <- readr::read_csv(est_path, show_col_types = FALSE) |>
      dplyr::filter(singletons == "included") |>
      dplyr::mutate(
        S_Chao2      = dplyr::if_else(is.na(chao2),     NA_integer_, as.integer(round(chao2))),
        Chao2_95_LCI = dplyr::if_else(is.na(chao2_lci), NA_integer_, as.integer(round(chao2_lci))),
        Chao2_95_UCI = dplyr::if_else(is.na(chao2_uci), NA_integer_, as.integer(round(chao2_uci)))
      ) |>
      dplyr::select(stratum, S_Chao2, Chao2_95_LCI, Chao2_95_UCI)
    cov_tbl <- dplyr::left_join(cov_tbl, est_tbl, by = "stratum")
  }
  cov_tbl |>
    dplyr::filter(!startsWith(stratum, "Canada"), n_sites >= 10) |>
    show_table(caption = paste(
      "Table S1. Per-ecozone sample completeness (Ĉ) for Canadian EcM fungal SH",
      "codes, from the georeferenced GlobalFungi site × SH incidence matrix",
      "(ecozones with ≥10 unique sites; named and unnamed SH types included).",
      "n_sites: unique 3-decimal-binned sites (~100 m); Q1/Q2: SH codes detected",
      "at exactly 1 or 2 sites; coverage_hat: Ĉ at observed site count. S_Chao2:",
      "asymptotic SH richness (iNEXT, incidence_freq, q = 0, singletons",
      "included); Chao2_95_LCI/UCI: bootstrap 95% CI; NA where n_sites < 30."))
} else {
  cat("`linnean_extrapolation_coverage.csv` not found — run 09_linnean.R first.")
}
```

## Table S2. Host species with recorded EcM associations

```{r table-s2-host-species}
elt_match_path <- file.path(paths$out_eltonian, "eltonian_host_matching.csv")
if (file.exists(elt_match_path) && file.exists(paths$host_species)) {
  host_tbl      <- readr::read_csv(paths$host_species, show_col_types = FALSE)
  elt_matching  <- readr::read_csv(elt_match_path, show_col_types = FALSE)

  # A host_clean string is a species-level identification only if it has exactly
  # two tokens and is neither an "sp."/"cf."/"aff." epithet nor an "unidentified"
  # genus token.
  is_species_level <- function(x) {
    toks <- strsplit(x, " ", fixed = TRUE)
    vapply(toks, function(p) {
      length(p) == 2L &&
        !grepl("^(sp\\.?|spp\\.?|cf\\.?|aff\\.?|indet\\.?)$", p[2], ignore.case = TRUE) &&
        !grepl("^(unidentified|indet|unknown)$", p[1], ignore.case = TRUE)
    }, logical(1))
  }

  recorded_raw     <- sort(unique(elt_matching$host_clean))
  recorded_species <- recorded_raw[is_species_level(recorded_raw)]
  host_intersection <- intersect(host_tbl$species, recorded_species)

  tibble::tibble(`Host species` = paste0("<em>", sort(host_intersection), "</em>")) |>
    show_table(caption = paste0(
      "Table S2. The ", length(host_intersection), " FungalRoot/BIEN native ",
      "Canadian EcM host species with ≥1 recorded, species-level host taxon in ",
      "Canadian GlobalFungi or GenBank EcM records (the intersection between the ",
      "native host list and the distinct species-level host taxa recorded across ",
      "both sources)."))
} else {
  cat("`eltonian_host_matching.csv` or the host-species list not found — run 19_eltonian.R first.")
}
```

## Figure S1. Global GlobalFungi sampling density

```{r fig-s1-density}
#| out-width: "100%"
if (file.exists(paths$fig_density_world)) {
  knitr::include_graphics(paths$fig_density_world)
} else {
  cat("`Figure-S1_gf_sampling_density_world.png` not found — run 12_wallacean_density_map.R.")
}
```

**Figure S1.** Sampling density of GlobalFungi samples in which at least one EcM
fungal taxon was detected. Each coloured pixel represents a single
100 × 100 km region.

## Figure S2. EcM sampling locations by ecozone

```{r fig-s2-ecozone-map}
#| out-width: "100%"
if (file.exists(paths$fig_ecozone_sampling)) {
  knitr::include_graphics(paths$fig_ecozone_sampling)
} else {
  cat("`Figure-S2_ecozone_sampling_map.png` not found — run 17_hutchinsonian.R.")
}
```

**Figure S2.** EcM fungal sampling locations (GlobalFungi and GenBank) projected
in Canada Albers Equal Area and overlaid on Canadian ecozone polygons (National
Ecological Framework for Canada).

## Figure S3. GBIF physical specimen records

```{r fig-s3-gbif}
#| out-width: "100%"
if (file.exists(paths$fig_gbif_specimens)) {
  knitr::include_graphics(paths$fig_gbif_specimens)
} else {
  cat("`Figure-S3_gbif_specimens.png` not found — run 20_sampling_maps.R.")
}
```

**Figure S3.** GBIF physical EcM specimen records in Canada, split by whether the
genus also has molecular sequence data in Canada. Blue circles: genera with
sequence data; orange triangles: genera without.

# References

Chamberlain S et al. (2024) *rgbif: Interface to the GBIF API.* R package.

Chao A (1984) Nonparametric estimation of the number of classes in a
population. *Scandinavian Journal of Statistics* 11: 265–270.

Chao A & Jost L (2012) Coverage-based rarefaction and extrapolation. *Ecology*
93: 2533–2547.

Dornelas M et al. (2018) BioTIME: A database of biodiversity time series for the
Anthropocene. *Global Ecology and Biogeography* 27: 760–786.

Hijmans RJ (2024) *terra: Spatial Data Analysis.* R package.

Hsieh TC, Ma KH & Chao A (2016) iNEXT: rarefaction and extrapolation of species
diversity. *Methods in Ecology and Evolution* 7: 1451–1456.

Maitner BS et al. (2018) The BIEN R package. *Methods in Ecology and Evolution*
9: 373–379.

Moulatlet GM et al. (2025) Climatic niches predict plant range size and
ecological dominance globally. *Proceedings of the National Academy of Sciences*
122: e2517585122.

Nilsson RH et al. (2019) The UNITE database for molecular identification of
fungi. *Nucleic Acids Research* 47: D259–D264.

Oksanen J et al. (2022) *vegan: Community Ecology Package.* R package.

Pebesma EJ (2018) Simple features for R. *The R Journal* 10: 439–446.

Pedersen TL (2024) *patchwork: The Composer of Plots.* R package.

Põlme S et al. (2020) FungalTraits. *Fungal Diversity* 105: 1–16.

R Core Team (2026) *R: A Language and Environment for Statistical Computing.*
R Foundation for Statistical Computing, v4.5.2.

Rognes T et al. (2016) VSEARCH: a versatile open-source tool for metagenomics.
*PeerJ* 4: e2584.

Soudzilovskaia NA et al. (2020) FungalRoot. *New Phytologist* 227: 955–966.

Tedersoo L et al. (2021) The Global Soil Mycobiome consortium dataset. *Fungal
Diversity* 111: 573–588.

Větrovský T et al. (2020) GlobalFungi. *Scientific Data* 7: 228.

Wickham H (2016) *ggplot2: Elegant Graphics for Data Analysis.* Springer.

Winter DJ (2017) rentrez: an R package for the NCBI eUtils API. *The R Journal*
9: 520–526.
