---
subtitle: "Supplemental materials (SM1) for: An assessment of biodiversity data shortfalls for ectomycorrhizal fungi in Canada"
author: "Jason Pither, with assistance from Claude (Sonnet 5 and Opus 4.8)"
bibliography: references.bib
csl: facets.csl  
link-citations: true
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    df-print: kable
execute:
  echo: false
  warning: false
  message: false
editor: source
---

```{=html}
<style>
/* Left-justify all table header cells (applies to knitr::kable tables and
   pandoc-rendered markdown pipe tables alike). */
th {
  text-align: left !important;
}
</style>
```

```{r setup}
#| include: false
# Source the shared pipeline setup: loads packages, the `paths` list, helper
# functions, and (if it exists) the primary EcM dataset `emf`.
source(here::here("scripts", "00_setup.R"))

# Small helper for rendering data-frame tables with a caption.
show_table <- function(x, caption = NULL, ...) {
  knitr::kable(x, caption = caption, format = "html", escape = FALSE, ...)
}

# Convenience: absolute path to a figure by file name.
fig <- function(name) here::here("figures", name)

# ---- Values used in figure captions and the data-overlap paragraph ----------
# Wired to the pipeline outputs so the caption/prose numbers track the data.
nf  <- function(x) format(round(as.numeric(x)), big.mark = ",", trim = TRUE)
p1  <- function(x) sprintf("%.1f", as.numeric(x))
.rd <- function(path) if (file.exists(path)) readr::read_csv(path, show_col_types = FALSE) else NULL

# GenBank total EcM records (overlap paragraph)
.gbt     <- .rd(file.path(paths$out_eltonian, "eltonian_genbank_tissue_tally_canada.csv"))
gb_total <- if (!is.null(.gbt)) .gbt$n[.gbt$category == "Total EcM fungal records"][1] else NA_real_

# World GlobalFungi samples actually plotted in Figure S2 (i.e. with usable
# coordinates), read from the same checkpoint the density map is built from.
# A small number of quality-filtered EcM samples worldwide lack coordinates
# and so cannot be mapped; using the mapped count keeps the caption exact.
.gws <- .rd(file.path(paths$checkpoints, "gf_global_ecm_sample_ids.csv"))
n_world_ecm <- if (!is.null(.gws)) nrow(.gws) else NA_real_
n_can_gf    <- dplyr::n_distinct(dplyr::filter(emf, source == "GlobalFungi")$sample_ID)

# Unique sampling locations by source (Fig S4)
.wl     <- .rd(file.path(paths$out_wallacean, "wallacean_location_summary.csv"))
loc_gf  <- if (!is.null(.wl)) .wl$n_unique_locations[.wl$dataset == "GlobalFungi"][1] else NA_real_
loc_gb  <- if (!is.null(.wl)) .wl$n_unique_locations[.wl$dataset == "GenBank"][1]     else NA_real_
loc_all <- if (!is.null(.wl)) .wl$n_unique_locations[.wl$dataset == "Combined"][1]    else NA_real_

# Sequencing depth discarded by EcM-genus filtering (Fig S3; same calc as 20_depth_discard.R)
.discard <- tryCatch({
  sm  <- readr::read_csv(paths$gf_meta_out, show_col_types = FALSE) |>
    dplyr::filter(sample_type == "soil") |> dplyr::distinct(sample_ID, ITS_total)
  eps <- emf |>
    dplyr::filter(source == "GlobalFungi", sample_ID %in% sm$sample_ID, !is.na(sh_code), abundance > 0) |>
    dplyr::group_by(sample_ID) |> dplyr::summarise(ecm = sum(abundance), .groups = "drop")
  d <- sm |> dplyr::left_join(eps, by = "sample_ID") |>
    dplyr::mutate(ecm = dplyr::coalesce(ecm, 0), pct = 100 * (1 - ecm / ITS_total)) |>
    dplyr::filter(ecm >= 1)
  list(n = nrow(d), q = stats::quantile(d$pct, c(.25, .5, .75), names = FALSE))
}, error = function(e) list(n = NA, q = c(NA, NA, NA)))

# Native host list: size and growth-form provenance (host-species section).
.hs <- .rd(paths$host_species)
n_host_native <- if (!is.null(.hs)) nrow(.hs) else NA_real_
.gfsrc <- if (!is.null(.hs) && "growth_form_source" %in% names(.hs))
  table(.hs$growth_form_source) else NULL
# Absence of a source means zero species came from it.
gfn <- function(k) if (!is.null(.gfsrc) && k %in% names(.gfsrc)) as.integer(.gfsrc[[k]]) else 0L

# BIEN2 range availability: host-list size vs species that actually contribute a
# modelled range to the richness rasters (one stack layer per contributing species).
n_stack <- tryCatch(terra::nlyr(terra::rast(paths$bien_species_stack)), error = function(e) NA_integer_)
n_norange <- if (!is.na(n_stack) && !is.na(n_host_native)) n_host_native - n_stack else NA_integer_

# Georeferenced GBIF specimen records plotted within Canada (Fig S5).
# Written by 19_sampling_maps.R as the exact point counts drawn on the map.
.gpc <- .rd(paths$gbif_plotted_counts)
.gbif <- if (!is.null(.gpc)) list(
  withseq = .gpc$n_plotted[.gpc$category == "Genera with sequence data in Canada"][1],
  noseq   = .gpc$n_plotted[.gpc$category == "Genera without sequence data in Canada"][1]
) else list(withseq = NA_real_, noseq = NA_real_)
```

# Detailed methods

## Software

We conducted analyses in R (v4.5.2; @rCoreTeamLanguageEnvironment2026). We used the following R packages: `sf`
(@pebesmaSimpleFeaturesR2018) for vector spatial operations; `terra` (@hijmansTerraSpatialData2025) for
rasters; `ggplot2` (@wickhamGgplot2ElegantGraphics2016) and `patchwork` (@pedersenPatchworkComposerPlots2024) for figures;
`rentrez`
(@winterRentrezPackageNCBI2017) for GenBank retrieval; `rgbif` (@chamberlainRgbifInterfaceGBIF2024) for GBIF
access; `BIEN` (@maitner_bien_2018) for plant range and native-status data;
`geodata` (@hijmansGeodataAccessGeographic2026) and `rnaturalearth`
(@massicotteRnaturalearthWorldMap2026) for spatial base layers; `httr`
(@wickhamHttrToolsWorking2026) for programmatic download of BIEN2 range shapefiles;
`NSR` (@maitnerNSRNativeStatus2024) for native-status resolution;
`GIFT` (@denelleGIFTAccessGlobal2024)
for the plant growth-form traits used to fill gaps in the BIEN trait coverage;
`readxl` for reading FungalRoot's supplementary workbook; and `data.table` for
column-selective reads of the multi-gigabyte GlobalFungi matrices.
We used `ITSx` v1.1.3 (@bengtsson-palmeImprovedSoftwareDetection2013), with its
`HMMER` 3.x backend, to extract ITS sub-regions from GenBank sequences; `vsearch` v2.30.4
(@rognesVSEARCHVersatileOpen2016) for sequence-similarity searching; and `awk`
for extraction from the GlobalFungi abundance matrix.

## Assembly of the Canadian EcM fungal dataset

### Data sources

We compiled EcM fungal records for Canada from two publicly accessible sequence
sources: version 5 of the GlobalFungi database (@vetrovskyGlobalFungiGlobalDatabase2020) and NCBI
GenBank (accessed 29 June 2026). GlobalFungi is a curated, purpose-built
database of community-survey amplicon data, consistently processed but
aggregated from studies using varied sequencing platforms and protocols, and
primarily derived from soil samples. GenBank is a general-purpose sequence
repository comprising records deposited from diverse studies and sampling
efforts. 

### UNITE reference database and SH assignment

We assigned fungal species hypothesis (SH) codes and taxonomy using the UNITE
general FASTA release (build 4 April 2024, UNITE v10.0, DOI
10.15156/BIO/2959332; dynamic clustering, full-length ITS; @nilsson_unite_2019).
This release provides one representative sequence per SH code, each with a
single unambiguous taxonomy string in its header. We used the same UNITE reference database for both the GlobalFungi taxonomy and the GenBank
`vsearch` assignment described below, so both sources reference an identical SH
definition.

### GlobalFungi records

We downloaded the GlobalFungi v5 sample metadata file and the combined
ITS1/ITS2 SH abundance matrix. The GlobalFungi dataset comes pre-populated with SH codes that are based on the same UNITE release described above, and a pipeline described at [globalfungi.com](https://globalfungi.com).  We retained Canadian samples
by filtering for country = Canada; barcoding region in {ITS2, ITSboth};
non-manipulated conditions (`manipulated == "NO"`); and exclusion of substrate
types irrelevant to EcM fungi (shoot, air, water, sediment). We pivoted the
resulting wide abundance matrix to long format and removed zero-abundance
records, then joined UNITE taxonomy by SH code. 

### GenBank records

We searched GenBank for fungal ITS sequences from Canada using the
query `"Canada"[Country] AND "Fungi"[Organism] AND ("internal transcribed
spacer"[All Fields] OR "ITS"[All Fields])`, with no date cutoff and no
lifestyle keyword, so the search returns all matching fungal ITS records
regardless of guild. We retrieved FASTA sequences and record metadata
(coordinates, collection date, host, isolation source) in batches through the
NCBI Entrez API (`rentrez`; @winterRentrezPackageNCBI2017). 

To make the GenBank component comparable to the
GlobalFungi component (which we restricted to samples with a `barcoding_region`
of ITS2 or ITSboth) we used ITSx v1.1.3 (@bengtsson-palmeImprovedSoftwareDetection2013; HMMER backend,
Fungi profile set) to detect and extract the ITS1 and ITS2 sub-regions from each
retrieved sequence. We excluded records with no detectable ITS2 or with an extracted ITS2
fragment shorter than 100 bp before
SH assignment. We recorded whether each record also carried a genuine standalone
ITS1 detection in an `its_region` column (`ITS2` or `ITSboth`), directly
analogous to GlobalFungi's `barcoding_region`. 

We matched every eligible record against UNITE on its extracted ITS2 fragment
alone. The `its_region` flag is
for provenance only and does not influence assignment. This differs from the GlobalFungi pipeline, which independently classified each ITS region. Our pipeline preserves a one-row-per-accession schema, so we do not achieve exact parity
with GlobalFungi at the level of individual sub-region observations. To gauge
whether discarding ITS1 measurably costs resolution, our pipeline
cross-tabulates each record's assignment outcome (below) against whether it
carried a standalone ITS1 detection (`genbank_its1_resolution_crosstab.csv`).

We used `vsearch` v2.30.4 (@rognesVSEARCHVersatileOpen2016; `usearch_global`) to compare sequences against
the UNITE reference at 98.5% identity, matching GlobalFungi's documented
SH-assignment criterion. We searched both strands exhaustively (`--strand both
--maxaccepts 0 --maxrejects 0`) and retained all top-scoring hits per query
(`--top_hits_only`) so that we could resolve ties.  Where a query's top hits spanned more than one UNITE reference, we resolved the
tie by taxonomic agreement. We compared named-species hits
by species name, and "dark" hits (UNITE placeholders ending `_sp`) by SH code,
following UNITE's own dark-taxon convention (@nilsson_unite_2019; @rybergNewLightNamesNaming2018). When all tied hits agreed at that level, we assigned the record that SH (`taxonomic_resolution = "sh"`). When they disagreed but shared a genus,
we retained the record at genus-only resolution (`taxonomic_resolution =
"genus"`, with `sh_code` and `species` set to `NA` and the genus preserved),
following the principle of climbing to the lowest unambiguous rank (@heegerLongReadDNAMetabarcoding2018). We treated only records whose tied hits disagreed at genus level as genuinely ambiguous and excluded them (logged to
`genbank_ambiguous_species_excluded.csv`). Downstream SH-richness counts
therefore treat genus-resolved records as carrying no SH.

We obtained the taxonomy for SH-resolved records directly from the matched SH
representative in the UNITE reference. We obtained the host taxon from the structured GenBank `host` qualifier
where available, and used a fallback regex extraction of binomial names from the 
`isolation_source` field. We preserve the raw extracted string as `host_taxon_raw` (see below). We also assigned each GenBank record a `canada_basis`
provenance category, recording the evidence for its inclusion (structured country
field, coordinates, both, or search-only).

We addressed inconsistencies in raw host-plant names across the source datasets
(the GenBank `host` qualifier and `isolation_source` field, and GlobalFungi's
`dominant_plant_species` and `other_plant_species` fields) using a single shared
normalization function, `canonicalize_host()` (`00_setup.R`), applied identically
wherever we compare a raw host string against a species list (GenBank assembly
here, and the Eltonian and Hutchinsonian analyses below). It reduces each raw
string to a standardized "Genus species" (or bare genus) form, or to a missing
value where the string cannot be resolved to a plausible Latin name. This step is
necessary because host names in both sources are free text rather than a
controlled vocabulary, so the same species may appear as a full binomial, a
misspelling, an abbreviation, a common name, or with inconsistent capitalization
or stray punctuation.

### Combining sources and spatial validation

We combined the GlobalFungi and GenBank datasets by row-binding. For all records with parsed coordinates we applied
a spatial-containment check against the GADM administrative boundary for Canada (see below),
buffered by 2 km (applied in the Canada Albers Equal Area Conic projection, so
the buffer distance is a true 2,000 m regardless of latitude), recording the
result in a logical `coord_in_canada` column (TRUE = validated inside the
buffered boundary; FALSE = coordinates present but outside; NA = no
coordinates); original coordinates are always preserved. The buffer absorbs GADM
coastline-simplification and topology artefacts, which would otherwise exclude a
handful of coastal and small-island records (e.g. Gulf of St. Lawrence) carrying
`country == "Canada"` metadata. All spatial analyses filter on `coord_in_canada ==
TRUE`, while non-spatial enumerations (taxon richness, record counts) include
all records, treating a GlobalFungi `country == Canada` label as authoritative
even where coordinates fall marginally outside the buffered Canada polygon.

### Assigning EcM guild status

We determined fungal EcM status at the genus level using FungalTraits (@polmeFungalTraitsUserfriendlyTraits2020): we retained records whose UNITE-assigned genus carries
`primary_lifestyle = ectomycorrhizal` in FungalTraits (genus names lower-cased
before matching), and carried two additional FungalTraits fields, EcM lineage
and exploration type, into the final dataset. The resulting primary analysis
dataset (`data_derived/emf_canada_em_only.csv`) has one row per
(sample × SH code) detection.

### Definition of a "site"

For analyses requiring georeferenced sampling units, we binned sample
coordinates to three decimal places of latitude and longitude (~100 m at
temperate latitudes), and refer to these bins as sampling "sites". This
collapses potential GPS-precision noise across repeat readings at one general location. We used site-binning for the ecozone sampling-coverage
counts; the Wallacean occurrence counts (below) are intentionally left unbinned
so that each distinct coordinate contributes an occurrence.

## Reference data preparation

### Spatial layers

We obtained Canadian national, provincial, and territorial boundaries from GADM
via `geodata`, a higher-resolution Canada boundary and major-lake polygons from
Natural Earth via `rnaturalearth`, and Canadian ecoregions (National Ecological
Framework for Canada) with their ecozone-name lookup. We reprojected layers to
WGS84 for coordinate operations and to the Canada Albers Equal Area Conic
projection for mapping, simplified them with `terra` (`simplifyGeom`) where
appropriate, and wrote them to `data_derived/spatial/`.

### FungalRoot host-determination rule

We determined EcM host status for plant species and genera from FungalRoot (@soudzilovskaiaFungalRootGlobalOnline2020) using two independent routes.

For the species-level route, we used Darwin Core Archive occurrence records from GBIF (DOI 10.15468/a7ujmj), joining the occurrence file to the "measurements" extension on the mycorrhiza-type field. We classified a species as EcM if FungalRoot carries at least one occurrence with an unambiguous EcM-positive label ("EcM", "EcM, no AM colonization", or "ErM,EcM"). Two further labels - "EcM, AM undetermined" and "EcM,AM" - do not on their own qualify a species as EcM-demonstrated, because they are too ambiguous to serve as positive evidence. However, records carrying these labels are not treated as evidence against EcM status either: a species with such a record is still classified as EcM-demonstrated if at least one other, unambiguous record for that species exists elsewhere in FungalRoot. We excluded `EcM, AM undetermined` because, in at least one case we identified (the fern *Dryopteris filix-mas*), it was contradicted by separate AM and non-mycorrhizal records for the same species, consistent with this label denoting an EcM report for which AM status was simply not surveyed, rather than a confirmed AM-negative result. We excluded "EcM,AM" (a dual-positive call, indicating both EcM and AM colonization were reported) as qualifying evidence at the species level because it was frequently the only EcM signal available for a species, with no unambiguous EcM record found anywhere else in that species' genus. Of an initial 698 candidate species that qualified under a broader rule, 118 species (spanning 20 genera, including *Acer*, *Allium*, *Fraxinus*, and *Juglans*) relied solely on this dual-positive label. Given the lack of any corroborating unambiguous evidence, we judged this label too weak on its own to support an EcM-demonstrated classification (but see below).

For the genus-level route, we used FungalRoot's own published genus-level recommendation, Supplementary Table S2 (@soudzilovskaiaFungalRootGlobalOnline2020), which assigns each genus a mycorrhizal-type call based on at least 67% consistency of species-level diagnosis. A genus qualifies as an EcM host genus if Table S2 assigns it "EcM" or "EcM-AM"; the latter captures genera with genuine, well-supported dual mycorrhizal associations (e.g., *Salix*, *Populus*, *Eucalyptus*) that the binary species-level occurrence rule cannot express. A handful of Table S2 rows carry non-standard free-text or apparent data-entry-error values (e.g., "species-specific: EcM or NM-AM"); these do not match either qualifying label, so we exclude them by default. The two routes are independent: a species/genus can qualify via its own direct occurrence record, via Table S2, or both, and a species need not have an occurrence record of its own to qualify if its genus is included via Table S2.

We acknowledge the species-level criterion is conservative. Some genera excluded under it, including *Acer*, *Fraxinus*, and *Juglans*, have published evidence of genuine EcM associations elsewhere in the literature (e.g., *Fraxinus* is listed among confirmed dual-mycorrhizal genera by @testeDualMycorrhizalPlants2020), evidence that is not captured by any unambiguous occurrence record in the FungalRoot GBIF extract used here. The source Table S2 does not resolve this gap: it independently classifies *Acer*, *Fraxinus*, and *Juglans* as AM, so the genus-level route adds no evidence for these three genera either. We chose not to supplement FungalRoot with case-by-case literature additions, to keep host determination reproducible from a single, clearly defined source; the cost of this choice is that our EcM host list under-counts true EcM hosts for a small number of genera whose FungalRoot evidence (occurrence- and genus-level alike) is exclusively ambiguous or absent.

### Native Canadian EcM host species

We identified native EcM host plant species occurring in Canada by querying BIEN
(@enquistBIENBiodiversityInformatics2026; @maitner_bien_2018) for the plant
species with Canadian distributions,
filtering to natives via the Native Species Resolver (NSR;
@boyleNativeSpeciesResolver2024), one of BIEN's modular data services
(@enquistBIENBiodiversityInformatics2026), which determines whether a species
is native or introduced to the political division of the observation by
comparing it against curated regional checklists. For Canadian determinations
NSR draws on the Database of Vascular Plants of Canada
(VASCAN; @brouilletDatabaseVascularPlants2025), which NSR last refreshed on
16 September 2024, and the World Checklist of Vascular Plants, delivered via
Plants of the World Online (WCVP v13; @govaertsWCVPWorldChecklist2024),
refreshed 21 May 2024. We queried NSR through its
R package (@maitnerNSRNativeStatus2024), and intersected the resulting
native flora against the EcM host determination described above. We retain a
native species if it qualifies under **either** route: its own species-level
occurrence record satisfies the species rule, **or** its genus qualifies under
FungalRoot's Table S2 genus rule. We combine the two routes with OR, as
implemented in `06_host_species.R` (Step 4). The resulting list
(`data_derived/ecm_native_canada_host_species.csv`) records which route(s) applied
for each species (`evidence_source`: "occurrence", "table_s2", or "both") and is
used in the Hutchinsonian and Eltonian analyses.

Combining the two routes with OR also mitigates a taxonomic-resolver limitation.
Because BIEN's occurrence-name scrubbing, the Native Status Resolver, and the
GBIF backbone underlying FungalRoot are independent taxonomic resolvers, a
genuinely native, EcM-demonstrated species can fail the species-level
intersection when no single name string clears both the native filter and the
species-level demonstrated filter simultaneously. The green/mountain alder complex
(*Alnus alnobetula* / *A. crispa* / *A. viridis*) is a case in point: the native
filter returns "Native" only under *Alnus alnobetula* while FungalRoot's
unambiguous species-level EcM evidence is filed under *Alnus crispa*, so the
species-level route alone would not match. Here the genus route rescues the
species — genus *Alnus* is called "EcM" in Table S2, so *Alnus alnobetula* is
retained (via `evidence_source == "table_s2"`). A residual limitation remains only
for species whose species-level name fails to match across resolvers **and** whose
genus is absent from Table S2; we flag this as a known limitation of unresolved
cross-database synonymy rather than applying a general synonym-harmonization step
(which could shift counts for many species).

We assigned each host species a growth form (tree, shrub, or herbaceous), which
we use for the tree / non-tree split reported in the main text and in Figure 4.
Growth form comes from a four-stage chain, applied in order and recorded per
species in the `growth_form_source` column of
`data_derived/ecm_native_canada_host_species.csv`: (i) the modal value in the
BIEN trait database (@enquistBIENBiodiversityInformatics2026; @maitner_bien_2018);
(ii) for species BIEN does not cover,
GIFT trait 1.2.1 (@denelleGIFTAccessGlobal2024, queried at
`agreement = 0.66`); (iii) for species neither database covers, the modal growth
form of that genus's other species in this host list; and (iv) a curated manual
override for any species still unresolved. Of the **`r n_host_native`** host
species, **`r gfn("bien")`** growth forms come from BIEN, **`r gfn("gift")`**
from GIFT, **`r gfn("congener")`** from the congener-modal step, and
**`r gfn("manual")`** from the manual override. We flag the congener-inferred
group explicitly because those values are inferred from congeners rather than
taken from a trait record for the species itself.

### BIEN2 host ranges and richness rasters

We obtained MaxEnt-based modelled range polygons (BIEN2; @moulatletGeneralLawsBiodiversity2025)
for the native Canadian EcM host species from biendata.org, downloaded
programmatically through the REST API, reprojected to WGS84, and clipped to the
Canada boundary. We rasterized these polygons onto a 0.5° geographic grid to
produce three layers: host species richness (count of host species with a range
polygon per cell); host data richness (count of host species with at least one
Canadian EcM molecular-detection record per cell); and host data proportion
(the ratio of the two).

Not every host species contributes a range. Of the **`r nf(n_host_native)`**
native Canadian EcM host species, **`r nf(n_stack)`** have a BIEN2 modelled
range that overlaps Canada and therefore appear in the richness rasters; the
remaining **`r nf(n_norange)`** do not, either because BIEN2 has no modelled
range for the species or because the modelled range does not intersect the
Canada boundary once clipped. The host-richness and host-proportion layers, and
Figure 4, are therefore based on the **`r nf(n_stack)`** species with usable
ranges rather than the full host list. This makes the mapped host coverage
conservative: a cell may hold EcM host species whose ranges BIEN2 does not
model. 

## Shortfall analyses

### Linnean shortfall

We quantified the Canadian EcM fungal inventory at three taxonomic levels:
unique UNITE SH codes, genera, and named species (excluding unresolved UNITE
entries ending in `_sp`). SH codes are the primary richness unit because they
provide a consistent operational taxonomic unit applicable to both sources, and
because many EcM taxa are not yet formally named. We compared observed Canadian
EcM genera to the global EcM genus inventory in FungalTraits, and quantified the
overlap between the GlobalFungi and GenBank components at the named-species SH
level using set operations (shared, GlobalFungi-only, GenBank-only). We
characterized the dataset "dark fraction" as the proportion of detected SH codes
lacking a species-level UNITE name, computed for the combined dataset and for
each source separately. This SH-level dark fraction is not directly comparable
to the OTU-level dark-diversity estimates of @vangalenBiogeographyConservationEarths2025, whose
denominators include OTUs matched only to genus level.

We also computed the average per-sample proportion of species-unassigned ("dark")
OTUs among van Galen et al.'s Canadian forest and tundra samples, using their
published sample-specific estimates (Figshare DOI 10.6084/m9.figshare.28830371) —
a table giving, for each GlobalFungi v5 sample in their global compilation, the
proportion of EcM OTUs assigned and unassigned to a named species, together with
sample-level metadata including country and biome. We filtered this table to
samples with country = Canada and biome in {forest, tundra} (the subset of their
global sampling most comparable to our own) and took the unweighted mean across
those samples. We report this in the main text alongside our own SH-level
dark-fraction estimate as a rough, non-equivalent point of reference rather than a
direct comparison: the two measures use different taxonomic units (OTU versus SH
code) and different underlying clustering/assignment pipelines.

### Wallacean shortfall

We characterized geographic sampling breadth for each EcM taxon at three levels
(SH code, genus, named species). For each taxon we counted unique sampling
occurrences as distinct 30 arc-second grid cells (~1 km²), and produced
frequency distributions of occurrences per taxon within Canada and globally. We
report the proportion of taxa known from a single occurrence, and from fewer
than 30 occurrences, both within Canada and globally, as straightforward
measures of the shortfall. The within-Canada and global occurrence counts are
drawn from the GlobalFungi dataset only, because it provides the structured
sampling frame required for meaningful occupancy comparison; the global counts
map each named Canadian species (or SH code) to all globally-known SH codes
carrying that identity and match them against the full global GlobalFungi v5 SH
abundance matrix under the same quality filters used for the Canadian data. As a
complementary view relevant to species distribution modelling, we tabulated how
many taxa meet common minimum-occurrence thresholds for modelling (30 occurrence records).

### Prestonian shortfall

We obtained BioTIME (@dornelasBioTIMEDatabaseBiodiversity2018) directly from the BioTIME data
portal (biotime.st-andrews.ac.uk) on 5 March 2026, downloading both the raw
biodiversity time-series data (as an RDS file) and the accompanying study
metadata (as a CSV) via the portal's direct-download endpoints. We searched
this database for
time-series records of EcM fungal communities in Canada, filtering to studies
with at least one Canadian location, at least two sampling time points, and an
organism type consistent with fungi, and cross-referencing matching studies
against the EcM dataset by taxonomy. We operationalize the Prestonian
shortfall as the percentage of named EcM fungal species present in Canada that
lack any abundance time-series data recorded anywhere on Earth.

### Darwinian shortfall

We assessed genome-sequence availability for Canadian EcM genera by
cross-referencing detected genera against the JGI MycoCosm fungal-genomics
portal. We obtained MycoCosm's organism list on 5 March 2026 from its public
"All Fungi" listing page (mycocosm.jgi.doe.gov/fungi/fungi.info.html, no login
required), which lists every fungal genome in the portal; at the time of data
collection the portal's automated bulk-export feature required a JGI login and
was not functioning, so we retrieved the listing table by selecting and
copying its contents directly from the page. Because the source page lists fungal genomes
exclusively, we applied no additional kingdom- or phylum-level filtering. We operationalize the
Darwinian shortfall as the percentage of named Canadian EcM fungal species
absent from the database.

### Raunkiæran shortfall

We quantified trait-data availability using FungalTraits (a genus-level
compilation) for all EcM genera detected in Canada. We assessed the six FungalTraits columns carrying
information meaningful for EcM fungi, in two classes: three
morphological/structural traits derivable from taxonomic description (fruitbody
type, hymenium type, mycorrhizal exploration type) and three
ecological/interaction traits (secondary lifestyle, endophytic-interaction
capability, host specificity). We excluded the remaining FungalTraits columns as
inapplicable to EcM fungi by lifestyle — those describing decay and saprotrophic
modes, lichen photobionts, plant pathogenicity, animal biotrophy, and aquatic
habitat. The seventh potentially relevant column, growth form, is "filamentous
mycelium" for all EcM genera, and we excluded it as uninformative. Coverage is the proportion of genus × trait cells with a
documented value, where "documented" means non-missing and not the literal
entry "unknown". Blanks in secondary lifestyle and endophytic-interaction
capability are ambiguous between "not assessed" and "trait absent", so coverage
for those two traits is a lower bound.

### Hutchinsonian shortfall

We addressed environmental coverage at two levels. 

For **climate-space coverage**, we characterized how completely Canada's climate
space has been sampled. We defined climate space using WorldClim 2.1
bioclimatic variables BIO1 (mean annual temperature) and BIO12 (mean annual
precipitation), aggregating the 30 arc-second rasters to ~10 arc-minutes and
masking to the Canada boundary, then binned the two-dimensional
temperature × precipitation space into a 50 × 50 grid of equal-width climate
zones. For each zone we computed coverage as the proportion of its Canadian grid
cells containing at least one EcM sample (each cell counted once), with a
coverage of zero denoting a climate present in Canada but never sampled.
We mapped coverage in both geographic and climate space, separately for
GlobalFungi samples and for the combined GlobalFungi + GenBank dataset, against
the available-climate frequency (the number of Canadian cells per zone). We
adapted the code producing this figure from @hebertBlitzGapNationwide2026.

For **ecozone coverage**,
we determined which of Canada's 15 terrestrial ecozones (National Ecological
Framework for Canada; [ecozones.ca](https://www.ecozones.ca)) contain at least one EcM sampling location, and report
the number and proportion of ecozones sampled at a range of site thresholds.

### Eltonian shortfall

We first assessed the **availability of host information** in the raw sample
metadata. For GlobalFungi, among Canadian samples carrying at least one EcM SH
code we tallied how many had an entry in the `dominant_plant_species` field, and
cross-tabulated the samples with host information by `sample_type` (soil vs.
root) (`18_eltonian.R`, Step A2b). For GenBank, we binned each Canadian EcM
record's `isolation_source` string into tissue categories (root/ectomycorrhiza,
soil/rhizosphere, other, or not recorded) using keyword matching (`18_eltonian.R`,
Step A2c). Note that this availability diagnostic keys on the single
`dominant_plant_species` field for GlobalFungi, whereas the fungus–host
association analysis below draws host identities from **both**
`dominant_plant_species` and `other_plant_species`; the two therefore measure
subtly different quantities (presence of a primary host label vs. any usable host
identity), so their sample counts need not coincide.

We then focused on EcM fungus–host plant associations, in two scopes. For the
**Canadian scope**, we extracted host information from the GenBank `host`
qualifier (with fallback from `isolation_source`) and from GlobalFungi's
`dominant_plant_species` and `other_plant_species` fields, restricted to root
samples. We canonicalized all host strings with the shared helper and
matched them against the native Canadian EcM host species list, which serves as
the denominator of potential interactions. We built fungus × host interaction
matrices at the SH-code, named-species, and genus levels, and computed the
proportion of Canadian EcM host species with at least one documented fungal
association, the fill rate of the full host × named-fungal-species matrix, and
the proportion of observed pairs supported by only a single record. For the
**global scope**, we queried the full GlobalFungi global metadata for root
samples matching Canadian host species, and queried GenBank globally for each
Canadian EcM genus (retrieving all matching records), to establish which
Canadian hosts and fungi have any documented association anywhere. To
contextualize the sampling gap relative to host diversity, we produced a
bivariate map classifying each 0.5° grid cell by EcM host species richness
versus the proportion of those host species with EcM molecular-detection data in
Canada, shown separately for tree and non-tree hosts (this bivariate map is
produced in `18_eltonian.R`).

### Extra analyses 

We compiled physical EcM specimen records from GBIF as a vouchered
complement to the sequence data: an occurrence download for Kingdom Fungi in
Canada (basis of record `PRESERVED_SPECIMEN` or `LIVING_SPECIMEN`; records with
geospatial issues excluded), filtered to species-rank records in EcM genera. For
the primary map (Figure S5), the "with-sequence" genera are those observed in our
own Canada-specific dataset (itself restricted to FungalTraits EcM genera), while
Table S4's "without-sequence" genera are drawn directly from the FungalTraits EcM
genus list (`09_linnean.R`). The download (retrieved 6 March 2026; 673,532 occurrence records
across 294 datasets before filtering) is archived at
<https://doi.org/10.15468/dl.92rns5> (download key `0017649-260226173443078`;
predicates country = CA, `HAS_GEOSPATIAL_ISSUE` = FALSE, taxonKey = 5,
`BASIS_OF_RECORD` in {`PRESERVED_SPECIMEN`, `LIVING_SPECIMEN`}, format
`SIMPLE_CSV`; licence CC BY-NC 4.0).

## Figure preparation

Every figure in the manuscript and in these supplemental materials is generated
directly by the analysis scripts, with one exception described below. Each
analysis script writes its figure in two versions: a white-background version,
which is the version reproduced in the manuscript and supplements, and an
otherwise identical version on a light grey background (`#F2F2F2`).

We used the grey-background versions to assemble Figure 5, which is a designed
conceptual composite rather than a computed figure. We laid out its six source
panels manually in Microsoft PowerPoint and added the annotations that appear in
it — the quadrant labels and the ellipses that link sampling strategies to the
shortfalls they would jointly address. Figure 5 is therefore the only figure that
cannot be regenerated by running the pipeline. Its source panels can be, and are
listed below.

```{r table-fig5-sources}
tibble::tibble(
  `Source panel (grey-background version)` = c(
    "Figure-01_sampling_map_grey.png",
    "Figure-03_climate_gap_grey.png",
    "Figure-04_host_bivariate_map_grey.png",
    "Figure-S1_dark_diversity_grey.png",
    "Figure-S4_ecozone_sampling_map_grey.png",
    "Figure-S5_gbif_specimens_grey.png"
  ),
  `Produced by` = c(
    "19_sampling_maps.R",
    "17_hutchinsonian.R",
    "18_eltonian.R",
    "10_dark_diversity.R",
    "17_hutchinsonian.R",
    "19_sampling_maps.R"
  )
) |>
  show_table(caption = paste0(
    "Source panels used to assemble Figure 5, and the analysis script that ",
    "produces each."
  ))
```

# Supplemental figures and tables

```{r table-s1-ecozone}
ez_path <- file.path(paths$out_hutchinsonian, "hutchinsonian_ecozone_sample_counts.csv")
if (file.exists(ez_path)) {
  ez_raw <- readr::read_csv(ez_path, show_col_types = FALSE)
  ez_tbl <- ez_raw |>
    dplyr::arrange(dplyr::desc(density_total)) |>
    dplyr::transmute(
      Ecozone                        = ecozone,
      `GlobalFungi (loc.)`           = gf_locations,
      `Total (loc.)`                 = total_locations,
      `Area (km²)`                   = round(area_km2),
      `Density (loc. / 10,000 km²)`  = round(density_total, 2)
    )
  ez_tbl |>
    show_table(caption = paste0(
      "<strong>Table S1.</strong> Unique EcM fungal sampling locations per Canadian ecozone ",
      "(National Ecological Framework), for the combined GlobalFungi + GenBank ",
      "dataset (n = ", nf(sum(ez_raw$total_locations)), " locations across ",
      nrow(ez_raw), " ecozones) and for GlobalFungi only (n = ",
      nf(sum(ez_raw$gf_locations)), " locations), with ecozone area and combined ",
      "sampling density. Points are assigned to ecozones by a within-polygon join ",
      "against the detailed (unclipped) ecoregion layer, with points falling just ",
      "offshore of the ecoregion coastline snapped to the nearest ecozone within ",
      "5 km; ecozone areas use the same polygon basis, so counts and areas are ",
      "consistent. Ordered by decreasing combined density."))
} else {
  cat("`hutchinsonian_ecozone_sample_counts.csv` not found — run 17_hutchinsonian.R first.")
}
```

```{r table-s2-host-species}
elt_match_path <- file.path(paths$out_eltonian, "eltonian_host_matching.csv")
if (file.exists(elt_match_path) && file.exists(paths$host_species)) {
  host_tbl      <- readr::read_csv(paths$host_species, show_col_types = FALSE)
  elt_matching  <- readr::read_csv(elt_match_path, show_col_types = FALSE)

  # A host_clean string is a species-level identification only if it has exactly
  # two tokens and is neither an "sp."/"cf."/"aff." epithet nor an "unidentified"
  # genus token.
  is_species_level <- function(x) {
    toks <- strsplit(x, " ", fixed = TRUE)
    vapply(toks, function(p) {
      length(p) == 2L &&
        !grepl("^(sp\\.?|spp\\.?|cf\\.?|aff\\.?|indet\\.?)$", p[2], ignore.case = TRUE) &&
        !grepl("^(unidentified|indet|unknown)$", p[1], ignore.case = TRUE)
    }, logical(1))
  }

  recorded_raw     <- sort(unique(elt_matching$host_clean))
  recorded_species <- recorded_raw[is_species_level(recorded_raw)]
  host_intersection <- intersect(host_tbl$species, recorded_species)

  tibble::tibble(`Host species` = paste0("<em>", sort(host_intersection), "</em>")) |>
    show_table(caption = paste0(
      "<strong>Table S2.</strong> The ", length(host_intersection), " FungalRoot/BIEN native ",
      "Canadian EcM host species with ≥1 recorded, species-level host taxon in ",
      "Canadian GlobalFungi or GenBank EcM records (the intersection between the ",
      "native host list and the distinct species-level host taxa recorded across ",
      "both sources)."))
} else {
  cat("`eltonian_host_matching.csv` or the host-species list not found — run 18_eltonian.R first.")
}
```


```{r table-s3-unmatched-hosts}
if (file.exists(elt_match_path) && file.exists(paths$host_species)) {
  unmatched <- setdiff(recorded_species, host_tbl$species)
  tibble::tibble(`Host name (metadata)` = paste0("<em>", sort(unmatched), "</em>")) |>
    show_table(caption = paste0(
      "<strong>Table S3.</strong> The ", length(unmatched), " species-level host names recorded in the ",
      "Canadian GlobalFungi or GenBank metadata that did not match the FungalRoot/BIEN ",
      "native EcM host species list (Table S2). These include ornamental or non-native ",
      "species, misspellings, and other problematic or truncated entries, and are ",
      "excluded from the host-interaction analyses."))
} else {
  cat("`eltonian_host_matching.csv` or the host-species list not found — run 18_eltonian.R first.")
}
```


```{r table-s3-gbif-genera}
if (file.exists(paths$gbif_ecm_nosequence)) {
  gbif_genera <- readr::read_csv(paths$gbif_ecm_nosequence, show_col_types = FALSE) |>
    dplyr::filter(!is.na(genus), genus != "") |>
    dplyr::group_by(Genus = genus) |>
    dplyr::summarise(
      `GBIF records` = dplyr::n(),
      `GBIF species` = dplyr::n_distinct(species[!is.na(species) & species != ""]),
      .groups = "drop") |>
    dplyr::arrange(dplyr::desc(`GBIF records`))
  gbif_genera |>
    show_table(caption = paste0(
      "<strong>Table S4.</strong> The ", nrow(gbif_genera), " EcM fungal genera (FungalTraits) ",
      "represented in the Canadian GBIF physical-specimen records (Figure S5) that ",
      "have no sequence data in Canada (not detected in the combined GlobalFungi + ",
      "GenBank dataset), and so could be targeted for future molecular sampling. ",
      "'GBIF records' and 'GBIF species' are the count of physical-specimen ",
      "occurrence records and distinct species names per genus in the GBIF download."))
} else {
  cat("`linnean_gbif_ecm_nosequence_canada.csv` not found — run 09_linnean.R first.")
}
```

```{r figS1-dark}
#| out-width: "100%"
#| fig-cap: '**Figure S1.** Modelled dark ("undescribed") EcM fungal diversity across Canada (percentage of EcM OTUs undescribed), clipped from the global geospatial dataset of @vangalenBiogeographyConservationEarths2025.'
if (file.exists(paths$fig_dark_diversity)) {
  knitr::include_graphics(paths$fig_dark_diversity)
} else {
  cat("`Figure-S1_dark_diversity.png` not found — run 10_dark_diversity.R.")
}
```

```{r figS2-density}
#| out-width: "100%"
#| fig-cap: !expr paste0("**Figure S2.** Sampling density of GlobalFungi samples in which at least one EcM fungal taxon was detected (n = ", nf(n_world_ecm), "; in Canada, n = ", nf(n_can_gf), " samples). Each coloured pixel represents a single 100 × 100 km region.")
if (file.exists(paths$fig_density_world)) {
  knitr::include_graphics(paths$fig_density_world)
} else {
  cat("`Figure-S2_gf_sampling_density_world.png` not found — run 12_wallacean_density_map.R.")
}
```

```{r figS3-depth}
#| out-width: "100%"
#| fig-cap: !expr 'paste0("**Figure S3.** Distribution of the percentage of each Canadian GlobalFungi soil sample''s total sequencing depth discarded by EcM-genus filtering, for every sample with at least one EcM-classified read (n = ", nf(.discard$n), " samples). Q1–median–Q3: ", p1(.discard$q[1]), "–", p1(.discard$q[2]), "–", p1(.discard$q[3]), "%. The vertical dashed line marks the median (", round(.discard$q[2]), "%).")'
if (file.exists(paths$fig_depth_discard)) {
  knitr::include_graphics(paths$fig_depth_discard)
} else {
  cat("`Figure-S3_depth_discard.png` not found — run 20_depth_discard.R.")
}
```

```{r figS4-ecozone-map}
#| out-width: "100%"
#| fig-cap: !expr 'paste0("**Figure S4.** EcM fungal sampling locations (GlobalFungi and GenBank) projected in Canada Albers Equal Area and overlaid on Canadian ecozone polygons (National Ecological Framework for Canada). Sampling unit: one distinct raw lat/lon coordinate per record (GlobalFungi: n = ", nf(loc_gf), "; GenBank: n = ", nf(loc_gb), "; combined: n = ", nf(loc_all), " unique locations). Blue circles: GlobalFungi; orange triangles: GenBank.")'
if (file.exists(paths$fig_ecozone_sampling)) {
  knitr::include_graphics(paths$fig_ecozone_sampling)
} else {
  cat("`Figure-S4_ecozone_sampling_map.png` not found — run 17_hutchinsonian.R.")
}
```

```{r figS5-gbif}
#| out-width: "100%"
#| fig-cap: !expr 'paste0("**Figure S5.** Occurrence records associated with physical EcM fungal specimen records in Canada, retrieved from GBIF on 6 March 2026 (`PRESERVED_SPECIMEN` and `LIVING_SPECIMEN`; unique occurrence records). Blue circles: occurrence records of EcM fungal genera (FungalTraits) with sequence data in Canada (combined GlobalFungi + GenBank; n = ", nf(.gbif$withseq), " plotted records). Orange triangles: occurrence records of EcM fungal genera without any sequence data in Canada (n = ", nf(.gbif$noseq), " plotted records). Plotted records are the georeferenced, distinct-coordinate occurrences within the Canada boundary, far fewer than the total specimen records because most GBIF records lack coordinates or share a coordinate. Source: GBIF Occurrence Download <https://doi.org/10.15468/dl.92rns5>.")'
if (file.exists(paths$fig_gbif_specimens)) {
  knitr::include_graphics(paths$fig_gbif_specimens)
} else {
  cat("`Figure-S5_gbif_specimens.png` not found — run 19_sampling_maps.R.")
}
```

# References {.unnumbered}

::: {#refs}
:::
